Modelling state-switching functional biologging data with hidden Markov models

### 2020-11-21 Report/Sections/discussion.tex
Address comments (still need to clean figures)

Current FDA literature addresses dependence between curves either with multilevel models \citep{Chen:2012,Di:2009}, which lack a time component, or with functional time series, which overlook the possibility that curves have several distinct ``types" \citep{Kokoszka:2018}. Our work addresses these issues and introduces a flexible framework to model functional time series data using HMMs.
We deal with temporal dependence between curves with distinct ``types" by using either an HMM or a CarHMM to model the curve sequence. We then view each individual curve as an HMM emission whose distribution is described by yet another fine-scale model. Here we use a CarHMM as the fine-scale model, but there are a wide range of possible fine-scale models, including a Poisson process or continuous time approach similar to that of \citet{Michelot:2019}. We also incorporate a moving-window transformation into the fine-scale model to deal with intricate dependence structures on short time scales. Together, the coarse- and fine-scale models make up a hierarchical structure which can account for simultaneous processes taking place at different time scales. Provided the construction is not overly complex, a hierarchical model created using our method can be both flexible and easy to fit using maximum likelihood estimation.

We demonstrate the usefulness of this framework using an ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a northern resident killer whale off the coast British Columbia, Canada. This gives a deeper understanding of the animal's tri-axial movement and thus its energy expenditure \citep{Gleiss:2011,Qasem:2012}, which is essential when discerning the survival of both northern resident killer whales and their prey \citep{Noren:2011}. This model is likely applicable to many other diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}. In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time series data in a variety of fields.

@@

Current FDA literature addresses dependence between curves either with multilevel models with random effects and no time component at all \citep{Chen:2012,Di:2009}, or with functional time series which model the curves as an evolving process \citep{Kokoszka:2018} which overlook the possibility that curves have several distinct ``types". However, many curve sequences exhibit both temporal dependence and a discrete number of types. One obvious example is the case study examined here, where one marine mammal dive serves a distinct purpose such as travelling, socializing, resting, or foraging \citep{Tennessen:2019a}. Another example is accelerometer data in school children, where bouts of activity can be associated with a behaviour such as running or walking. \citep{Morris:2007}. In this work, we deal with both temporal dependence between curves and distinct curve ``types" by using either an HMM or CarHMM. To help select the appropriate model between these two, lag plots can be used to test if there is significant auto-correlation between curves.

>>>>

Current Functional Data Analysis literature addresses dependence between curves either with multilevel models \citep{Chen:2012,Di:2009}, which lack a time component, or with functional time series, which overlook the possibility that curves have several distinct ``types" \citep{Kokoszka:2018}. Our work addresses these issues and introduces a flexible framework to model functional time series data using HMMs.
We deal with temporal dependence between curves with distinct ``types" by using either an HMM or a CarHMM to model the curve sequence. We then view each individual curve as an HMM emission whose distribution is described by yet another fine-scale model. Here we use a CarHMM as the fine-scale model, but there are a wide range of possible fine-scale models, including a Poisson process or continuous time approach similar to that of \citet{Michelot:2019}. We also incorporate a moving-window transformation into the fine-scale model to deal with intricate dependence structures on short time scales. Together, the coarse- and fine-scale models make up a hierarchical structure which can account for simultaneous processes taking place at different time scales. Provided the construction is not overly complex, a hierarchical model created using our method can be both flexible and easy to fit using maximum likelihood estimation.

We demonstrate the usefulness of this framework using a biomechanical/ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a northern resident killer whale in Queen Charlotte Sound off the coast British Columbia, Canada. This gives a deeper understanding of a killer whale's tri-axial movement and thus its behaviour and energy expenditure \citep{Gleiss:2011,Qasem:2012}, both of which are important for understanding the foraging ecology and nutritional status of the Northern Resident killer whales \citep{Noren:2011}. This model is applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}. In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time series data in a variety of fields.

@@

Current FDA literature addresses dependence between curves either with multilevel models with random effects and no time component at all \citep{Chen:2012,Di:2009}, or with functional time series which model the curves as an evolving process \citep{Kokoszka:2018} which overlook the possibility that curves have several distinct ``types". However, many curve sequences exhibit both temporal dependence and a discrete number of types. One obvious example is the case study of killer whale behaviour examined here, where one dive can serve different purposes such as travelling, socializing, resting, or foraging \citep{Tennessen:2019a}. Another example is accelerometer data in school children, where bouts of activity can be associated with a behaviour such as running or walking. \citep{Morris:2007}. In this work, we deal with both temporal dependence between curves and distinct curve ``types" by using either an HMM or CarHMM. To help select the appropriate model between these two, lag plots can be used to test if there is significant auto-correlation between curves.

### 2020-11-21 Report/Sections/results.tex
Address comments (still need to clean figures)

To illustrate the process of constructing a model using these HMM building blocks, we investigate the dive behaviour of a northern resident killer whale off the coast of British Columbia, Canada, and construct several candidate models to categorize and describe its diving behaviour.

Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}. HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g. \citet{DeRuiter:2017} and \citet{Isojunno:2017}). For killer whales, we are interested in categorizing different diving behaviours to identify potential foraging dives. Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller prey \citep{Ford:2009}. Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook \citep{Williams:2009,Noren:2011,Wright:2017}. It is therefore of interest to determine the energetic expenditure of killer whales hunting for Chinook as well as the frequency at which they catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}. Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the behavioural state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of a killer whale.

@@

The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm PDT and consist of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz using Customizable Animal Solutions Tracking (CATs) tags (model number \#\#\#). The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left). Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. Tagging the killer whale caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored. We also ignore all dives between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm due to a malfunction in the tagging technology. To process the data, we smooth the accelerometer and depth data by taking a moving average within a window of $1/10^{th}$ of a second. We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, the vector of observed coarse-scale observations $y = \left\{y_1,\ldots,y_{267}\right\}$ is a time series of dive durations in seconds ($s$). For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots \right\}$, which is a vector of the within-dive acceleration data in meters per second squared ($m/s^2$). Naturally, the collection of all acceleration data is $y^* = \left\{y^*_1,\ldots,y^*_{267}\right\}$.

@@

Defining a suitable model for these data involves selecting an appropriate model structure, emission distributions, and the correct number of hidden states for both the coarse and fine scale observations.

We first select an appropriate model structure for the coarse-scale observations of dive duration.
%
We do not use information criteria to select the number of dive types $N$ since these metrics tend to overestimate the number of states in biological processes \citep{Pohle:2017}. If the hidden states are well-separated, a lag plot should reveal $N$ distinct patterns, where each pattern corresponds to one dive type \citep{Lawler:2019}. This is unfortunately not the case for our killer whale data, but we still consult lag plots together with visual inspection to choose $N = 2$ dive types. The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).
%
Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Section 1.1 of the supplementary material), and visual inspection shows no obvious intricate structure. Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor a moving-window transformation is called for.
%
Given that dive $t$ is of type $i$, we assume that the dive duration $Y_t$ follows a Gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$:
%
$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
%
This is consistent with previous studies, including that of \citet{Barajas:2017}. 
%

We then select a model corresponding to the fine-scale observations of acceleration. Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$. Although $N^*$ is selected heuristically, we test the validity of this model using techniques in Section \ref{subsec:model_validation}.
%
In contrast to the coarse-scale observations, the fine-scale acceleration data exhibit significant sinusoidal behaviour. Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with window size $h=100$ (two seconds) and maximum frequency $\tilde{\omega}=10$ (5 Hz). 
%We select a window size of two seconds since it appears be the smallest window size that captures at least one full period of each ``wiggle" in the acceleration data. 
We then have that $\z_{t,\tilde t^*} = \left\{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\right\}$, where $\zone_{t,\tilde t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window. Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material). Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.

@@

We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation. Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model. The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. In addition to the CarHHMM-DFT, we consider three similar variations for comparison. As in the full model, all components of $\Z_{t,\tilde t^*}$ are assumed to be independent when conditioned on dive type and subdive state for each of the following models:

@@

Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes no auto-correlation within the fine-scale observations, the CarHHMM does not incorporate the ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.

@@

The coarse-scale emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, travelling, and searching for prey.
On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as hunting and socialization \citep{Tennessen:2019b}. However, no dive in this data set has a maximum depth deeper than 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100 meters. It is therefore unlikely that any of the dives used in this study are successful foraging dives.

The three subdive behaviours have distinct average ``wiggliness" values, as the means corresponding to $\Ztwo_{t,\tilde t^*}$ are separated by an order of magnitude for each subdive state (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest variance corresponding to $\Zone_{t,\tilde t^*}$. It also has the highest auto-correlation in $\Zone_{t,\tilde t^*}$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1. The auto-correlation of acceleration is also slightly lower than in subdive state 1. We therefore hypothesize that subdive state 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $\Ztwo_{t,\tilde t^*}$ and variance of $\Zone_{t,\tilde t^*}$ in subdive state 3 are both much higher than in the other two states. The auto-correlation of $\Zone_{t,\tilde t^*}$ is also much lower. This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix} \text{ and }$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}$$
%
for the transitions between dives. Within dives, we have
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
%
for dive types 1 and 2.
The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that about 79\% of dives are short dives of type 1. The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive. The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is in the less active subdive state 1 40\% of the time when in a dive of type 1 but only 9\% of the time when in a dive of type 2. Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives. Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.

@@

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}. Each observed dive duration is weighted by the decoded probability that it corresponds to a particular dive type. This results in two histograms - one corresponding to dive type 1 and another corresponding to dive type 2. Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT. See Sections 1.4 through 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $\z_{t,\tilde t^*}$. Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions. In particular, $\Ztwo_{t,\tilde t^*}$ is right-skewed and $\Zone_{t,\tilde t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

@@

Finally, the CarHMM-DFT, which lacks a hierarchical structure, produces similar fine-scale parameter estimates and state estimates compared to those of the CarHHMM-DFT. However, its lack of hierarchical structure means that it fails to differentiate between short and long dives. This model therefore does not give any information regarding the dive-level Markov chain or the relationship between the dive and subdive levels. In particular, the CarHMM-DFT does not indicate that the whale is more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives. See Section 1 of the supplementary material for a more complete set of results for each of the candidate models.

>>>>

To illustrate the process of constructing a model using these building blocks, we analyze the dive behaviour of a northern resident killer whale in Queen Charlotte Sound off the coast of British Columbia, Canada, and construct several candidate models to categorize and describe its diving behaviour.

Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}. HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g. \citet{DeRuiter:2017} and \citet{Isojunno:2017}). For killer whales, we are interested in categorizing different diving behaviours to identify potential foraging dives. Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller types of salmon \citep{Ford:2009}. Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook \citep{Williams:2009,Noren:2011,Wright:2017}. 
%It is therefore of interest to determine the energetic expenditure of killer whales hunting for Chinook as well as the frequency at which they catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}. 
Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of a killer whale.

@@

The data we use were collected on September 2, 2019 from 12:49 pm to 6:06 pm PDT and consist of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz using CATs biologger (Customizable Animal Tracking Solutions, www.cats.is). The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left). Tri-axial acceleration readings are common in these types of tags and are often used to infer animal behaviour such as foraging \citep{Cade:2017,Fehlmann:2017,Wright:2017}. The act of attaching and detaching the tag caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations taken during these time periods are ignored. There were also periods of time when the tag failed to record observations, resulting in data gaps between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm. To process the data, we smooth the accelerometer and time-depth recorder data by taking a moving average within a window of $1/10^{th}$ of a second. We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, the vector of observed coarse-scale observations $y = \left\{y_1,\ldots,y_{267}\right\}$ is a time series of dive durations in seconds. For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots \right\}$, which is a vector of the within-dive acceleration data in units of meters per second squared. Naturally, the collection of all acceleration data is denoted as $y^* = \left\{y^*_1,\ldots,y^*_{267}\right\}$.

@@

Defining a suitable model to describe this killer whale kinematic data set involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse and fine scale observations.

We do not use information criteria to select the number of dive types $N$ since these metrics tend to overestimate the number of states in biological processes \citep{Pohle:2017}. We instead plot the duration of each dive versus the duration of the dive preceding it (i.e., $y_t$ vs $y_{t-1}$ for $t \geq 2$). This type of visualization is known as a lag plot. If the hidden states are well-separated, a lag plot should reveal $N$ distinct patterns, where each pattern corresponds to one dive type \citep{Lawler:2019}. This is unfortunately not the case for our killer whale data, as there is one cluster of data centred at approximately $30$ seconds. However, long dives appear to be characterized by bouts of less ``wiggly" behaviour in the acceleration data compared to shorter dives, so we choose $N = 2$ to differentiate these dive types. The absence of a more principled method to select $N$ highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).
%
Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Section 1.1 of the supplementary material), and visual inspection shows no obvious intricate structure. Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor a moving-window transformation is called for.
%
Given that dive $t$ is of type $i$, we assume that the dive duration $Y_t$ follows a Gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$:
%
$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
%
This is consistent with previous studies, including that of \citet{Barajas:2017}. 
%

We then select a model corresponding to the fine-scale observations of acceleration. Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$. Although $N^*$ is selected heuristically, we test the validity of this model in Section \ref{subsec:model_validation}.
%
In contrast to the coarse-scale observations, the fine-scale acceleration data exhibit significant sinusoidal behaviour. Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with window size $h=100$ (two seconds) and maximum frequency $\tilde{\omega}=10$ (5 Hz). 
%We select a window size of two seconds since it appears be the smallest window size that captures at least one full period of each ``wiggle" in the acceleration data. 
We then have that $\z_{t,\tilde t^*} = \left\{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\right\}$, where $\zone_{t,\tilde t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window. Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material). Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.

@@

We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation (see Figure \ref{fig:CarHHMM-DFT}). The likelihood of this model is easily calculated using the forward algorithm and can be maximized with respect to the parameters above (see appendix for details). In addition to the CarHHMM-DFT, we consider three similar variations for comparison. As in the full model, all components of $\Z_{t,\tilde t^*}$ are assumed to be independent when conditioned on dive type and subdive state for each of the following models:

@@

Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes there is no auto-correlation between fine-scale observations, the CarHHMM does not incorporate the ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.

@@

The coarse-scale emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives which likely reflects resting, travelling, and (to a lesser extent) searching for prey.
On the other hand, dive type 2 is longer, deeper, and may be associated with behaviours such as hunting and socialization \citep{Tennessen:2019b}. No dive in this data set has a maximum depth deeper than 30 meters, and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters. However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and significant numbers of fish have been observed to be caught near the surface (Fortune and Trites, unpublished data). It is therefore unclear whether any of the dives in this study are successful foraging dives.

The means of the ``wiggliness" $\Ztwo_{t,\tilde t^*}$ are each separated by an order of magnitude between subdive states (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest variance corresponding to $\Zone_{t,\tilde t^*}$. It also has the highest auto-correlation in $\Zone_{t,\tilde t^*}$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1. The auto-correlation of acceleration is also slightly lower than in subdive state 1. We therefore hypothesize that subdive state 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $\Ztwo_{t,\tilde t^*}$ and variance of $\Zone_{t,\tilde t^*}$ in subdive state 3 are both much higher than in the other two states. The auto-correlation of $\Zone_{t,\tilde t^*}$ is also much lower. This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix} \text{ and }$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}$$
%
for the transitions between dives. Within dives, we have
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
%
for dive types 1 and 2.
The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that about 79\% of dives are short dives of type 1. The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive. The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is in the less active subdive state 1 40\% of the time during a dive of type 1 but only 9\% of the time during a dive of type 2. Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives. Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.

@@

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}. Each observed dive duration is weighted by the decoded probability that it corresponds to a particular dive type. This results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2. Each histogram is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT. Analogous histograms corresponding to the fine-scale observations $\z_{t,\tilde t^*}$ are contained in Sections 1.4 through 1.6 of the supplementary material. Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions. In particular, $\Ztwo_{t,\tilde t^*}$ is right-skewed and $\Zone_{t,\tilde t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

@@

Finally, the CarHMM-DFT, which lacks a hierarchical structure, produces similar fine-scale parameter estimates and state estimates compared to those of the CarHHMM-DFT. However, its lack of hierarchical structure means that it fails to differentiate between short and long dives. This model therefore does not give any information regarding the dive-level Markov chain or the relationship between the dive and subdive levels. In particular, the CarHMM-DFT does not indicate that the whale is more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives. A more complete set of results for each of the candidate models is in Section 1 of the supplementary material.
